---
title: "HW5 - Make Data Move"
author: "Astrid Rybner"
date: "05/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Explore the recent global developments with R
Today, you will load a filtered gapminder dataset -  with a subset of data on global development from 1952 - 2007 in increments of 5 years - to capture the period between the Second World War and the Global Financial Crisis. 

**Your task: Explore the data and visualise it in both static and animated ways, providing answers and solutions to 7 questions/tasks below.**

## Get the necessary packages
First, start with installing the relevant packages 'tidyverse', 'gganimate', and 'gapminder'.

```{r libraries, echo = FALSE}
library(tidyverse)
library(gganimate)
library(gapminder)

```

## Look at the data
First, see which specific years are actually represented in the dataset and what variables are being recorded for each country. Note that when you run the cell below, Rmarkdown will give you two results - one for each line - that you can flip between.
```{r}
unique(gapminder$year)
head(gapminder)

gapminder <- as.data.frame(gapminder)

```
The dataset contains information on each country in the sampled year, its continent, life expectancy, population, and GDP per capita.

Let's plot all the countries in 1952.
```{r 1957}
theme_set(theme_bw())  # set theme to white background for better visibility

ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() 

```
We see an interesting spread with an outlier to the right. Answer the following questions, please:

Q1. Why does it make sense to have a log10 scale on x axis?

#Answer Q1
As the differences in gdp are this big, you would not be able to see the richest countries within the plot, if you did not 'squeeze' the richest countries closer to the poorer one with this transformation of the x-axis. 

Q2. What country is the richest in 1952 (far right on x axis)? 
```{r}

gapminder %>% filter(year==1952) %>% arrange(desc(gdpPercap))

```
#Answer Q2
From the above code you can see, that the maximum gdp in 1952 was 108382.35 in the country Kuwait. 



You can generate a similar plot for 2007 and compare the differences
```{r 2007}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() 
```
The black bubbles are a bit hard to read, the comparison would be easier with a bit more visual differentiation.

Q3. Can you differentiate the continents by color and fix the axis labels?

#Answer Q3
You can differentiate between the continents by adding a color argument in the aestetics ie. color=continent.
You add labels to the x- and y-axis with the labs function, in which you specify, what you want to call the x- and y-axis respectively. You can also specify the title of the legends in the labs() function. The size= argument names the size lab and the color= argument defines what the color lab is called. If you eg. had 'shape' as a  facet in your plot, you could say shape="title" to name that argument. 
```{r}

ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop, color=continent)) + #color by continent
  geom_point() +
  scale_x_log10()+labs(x="GDP per capita", y="Life expentancy", colour="Continent", size="Population size") #adding axis labels

```


Q4. What are the five richest countries in the world in 2007?
```{r}
gapminder %>% filter(year==2007) %>% arrange(desc(gdpPercap)) %>% head(n=5)

```
#Answer Q4
The five richest countries in 2007 were Norway, Kuwait, Singapore, United States and Ireland. 



##  Make it move!

The comparison would be easier if we had the two graphs together, animated. We have a lovely tool in R to do this: the `gganimate` package. And there are two ways of animating the gapminder ggplot.

### Option 1: Animate using transition_states() 

The first step is to create the object-to-be-animated
```{r anim1 start}

anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10()  # convert x to log scale
anim

```

This plot collates all the points across time. The next step is to split it into years and animate it. This may take some time, depending on the processing power of your computer (and other things you are asking it to do). Beware that the animation might appear in the 'Viewer' pane, not in this rmd preview. You need to knit the document to get the viz inside an html file.

```{r anim1}
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1)

```
Notice how the animation moves jerkily, 'jumping' from one year to the next 12 times in total. This is a bit clunky, which is why it's good we have another option. 


### Option 2 Animate using transition_time()
This option smoothes the transition between different 'frames', because it interpolates and adds transitional years where there are gaps in the timeseries data.

```{r anim2}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() + # convert x to log scale
  transition_time(year)
anim2

```

The much smoother movement in Option 2 will be much more noticeable if you add a title to the chart, that will page through the years corresponding to each frame.


Q5 Can you add a title to one or both of the animations above that will change in sync with the animation? [hint: search labeling for transition_states() and transition_time() functions respectively]

```{r}
anim3 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color=continent)) +
  geom_point() +
  scale_x_log10() + # convert x to log scale
transition_states(year, transition_length = 1, state_length = 1)+labs(title = 'Year: {closest_state}')

anim3

```
#Answer Q5
In the above you define what one 'transition' is (here it is 1 step in the year variable) and you use these defined steps (in states) to make a title. 
I.e. you tell it that the title should be the closest 'state' at that point in the animation. 




Q6 Can you made the axes' labels and units more readable? Consider expanding the abreviated lables as well as the scientific notation in the legend and x axis to whole numbers.[hint:search disabling scientific notation]
```{r}
options(scipen=999) #setting a value for how likely it is that scientific notation is triggered

anim4 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color=continent)) +
  geom_point() +
  scale_x_log10() + # convert x to log scale
transition_states(year, transition_length = 1, state_length = 1)+labs(title = 'Year: {closest_state}', x="GDP per capita ($) on a log scale", y="Life expentancy (years)", colour="Continent", size="Population size")
anim4

```
#Answer Q6
Yes, with the options function. With the options function you set a value for how likely it is that scientific notation (with the e's) is triggered. If you set it to a large positive value then there is a very low chance that scientific notation is triggered. 




Q7 Come up with a question you want to answer using the gapminder data and write it down. Then, create a data visualisation that answers the question and explain how your visualization answers the question. (Example: you wish to see what was mean life expectancy across the continents in the year you were born versus your parents' birth years). [hint: if you wish to have more data than is in the filtered gapminder, you can load either the `gapminder_unfiltered` dataset and download more at https://www.gapminder.org/data/ ]


#Answer Q7 
Question: How was the development in GDP per capita in the two nordic countries Nowrway and Denmark in the period 1952-2007?

Here I use the cowplot package to plot the two plots next to each other and the viridis package to get some pretty colors on it. 

```{r}
#install.packages("cowplot") # package for plotting to plots in one
#install.packages("viridis")  # package for coloring
library(viridis) 
library(cowplot)
summary(gapminder)#to get summary stats for each collumn in the gapminder data

#using the theme() package to make pretty plots

plot_1 <- gapminder %>%  filter(country=="Denmark") %>% ggplot(., aes(x=year, y=gdpPercap)) + geom_point(aes(color=gdpPercap), show.legend = F)+labs(x="Year", y="GDP per capita", title="Economic development in Denmark", subtitle="1952-2007") + theme(plot.title = element_text(size=10), plot.subtitle=element_text(size=8), axis.title = element_text(size=8)) + scale_color_viridis(option = "D")+theme_minimal()



plot_2 <- gapminder %>%  filter(country=="Norway") %>% ggplot(., aes(x=year, y=gdpPercap)) + geom_point(aes(color=gdpPercap), show.legend = F)+labs(x="Year", y="GDP per capita", title="Economic development in Norway", subtitle="1952-2007") + theme(plot.title = element_text(size=10), plot.subtitle=element_text(size=8), axis.title = element_text(size=8)) + scale_color_viridis(option = "D")+theme_minimal()

plot_grid(plot_1, plot_2, labels="AUTO")

getwd()

```







